---
title: "p8105_hw2_yf2605"
author: "Yi_Fang"
date: "`r Sys.Date()`"
output: github_document
---

# Libraries

```{r}
library(tidyverse)
library(readxl)
```

# Problem 1 

## Read and clean the data

```{r}
entrance_exit_raw = 
  read_csv(file = 'data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv',
           col_types = cols(Route8=col_character(), 
                            Route9=col_character(),
                            Route10=col_character(),
                            Route11=col_character(),
                            .default = col_guess()))
entrance_exit_cleaned = 
  janitor::clean_names(entrance_exit_raw) %>% 
  select(line, station_name, station_latitude, 
         station_longitude, starts_with("route"), 
         entry, exit_only, vending, entrance_type, ada) %>% 
  mutate(entry = recode(entry, "YES" = TRUE, "NO" = FALSE))
```

## Brief description

The original dataset contains `r ncol(entrance_exit_raw)` variables: `r colnames(entrance_exit_raw)`. 

Data cleaning steps: 

*   cleaned variable names.
*   retained variables: `r colnames(entrance_exit_cleaned)`. 
*   converted variable `entry` from character to a logical variable, YES-->TRUE, NO-->FALSE.

The size of the resulting dataset is (row x column): `r nrow(entrance_exit_cleaned)` x `r ncol(entrance_exit_cleaned)`. 

These datasets are not tidy.

## Questions

*   How many distinct stations are there?

```{r}
stations = 
  distinct(entrance_exit_cleaned, line, station_name, .keep_all = TRUE)
```

There are `r nrow(stations)` distinct stations.

*   How many stations are ADA compliant?

```{r}
ada_sations = 
  filter(stations, ada == TRUE)
```

`r nrow(ada_sations)` stations are ADA compliant.

*   What proportion of station entrances / exits without vending allow entrance?

```{r}
no_vending_entrance_exit = 
  filter(entrance_exit_cleaned, vending == 'NO')

no_vending_entrance = 
  filter(no_vending_entrance_exit, entry == TRUE)
```

`r round(100 * nrow(no_vending_entrance) / nrow(no_vending_entrance_exit), 2)` % of station entrances / exits without vending allow entrance

*   Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?

```{r}
entrance_exit_tidy = 
  pivot_longer(
    entrance_exit_cleaned, 
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name")

a_train_stations = 
  filter(entrance_exit_tidy, route_name == 'A') %>% 
  distinct(line, station_name, .keep_all = TRUE)

a_train_ada_stations = 
  filter(a_train_stations, ada == TRUE) %>% 
  distinct(line, station_name)
  
```

`r nrow(a_train_stations)` distinct stations serve the A train.

Of the stations that serve the A train, `r nrow(a_train_ada_stations)` are ADA compliant.

# Problem 2

## Read and clean the Mr. Trash Wheel sheet

*   read data entries, excluded non-data area and the grand total row.
*   omit rows that do not include dumpster-specific data.
*   round the number of sports balls to the nearest integer and converts the result to an integer variable.

```{r}
mr_trash_wheel = 
  read_excel(path = 'data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx',
             sheet = 'Mr. Trash Wheel',
             range = "A2:N534") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(sports_balls = as.integer(round(sports_balls)))
```

## Read and clean the Professor Trash Wheel sheet

*   read data entries, excluded non-data area and the grand total row.
*   omit rows that do not include dumpster-specific data.
*   round the number of sports balls to the nearest integer and converts the result to an integer variable.

```{r}
professor_trash_wheel = 
  read_excel(path = 'data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx',
             sheet = 'Professor Trash Wheel',
             range = "A2:N116") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(sports_balls = as.integer(round(sports_balls)))
```

## Combine Mr. Trash Wheel and Professor Trash Wheel

*   add variable trash_wheel_type to keep track of Trash Wheel Types
*   stack mr_trash_wheel and professor_trash_wheel together

```{r}
mr_trash_wheel = 
  mutate(mr_trash_wheel, trash_wheel_type = 'Mr')

professor_trash_wheel = 
  mutate(professor_trash_wheel, trash_wheel_type = 'Professor')

trash_wheels = 
  rbind(mr_trash_wheel, professor_trash_wheel)
```

## Brief description

The dataset contains `r ncol(trash_wheels)` variables and `r nrow(trash_wheels)` records,  `r nrow(mr_trash_wheel)` records are from Mr. Trash Wheel and `r nrow(professor_trash_wheel)` records are from professor_trash_wheel.

The variables include:
`r colnames(trash_wheels)`

The the total weight of trash collected by Professor Trash Wheel is `r sum(professor_trash_wheel$weight_tons)` tons
The total number of sports balls collected by Mr. Trash Wheel in 2020 is `r sum(filter(mr_trash_wheel, year == 2020)$sports_balls)`.

# Problem 3

## clean the data in pols-month.csv

*   Use separate() to break up the variable mon into integer variables year, month, and day.
*   replace month number with month name.
*   create a president variable taking values gop and dem.
*   remove variables: prez_dem prez_gop day.
*   arrange according to year and month.
*   make sure year and month are the leading columns.

```{r}
pols_month = 
  read_csv(file = 'data/fivethirtyeight_datasets/pols-month.csv') %>% 
  janitor::clean_names() %>% 
  separate(mon, c('year', 'month', 'day'), convert = TRUE) %>% 
  mutate(month = month.name[month],
         president = if_else(prez_dem == 1, 'dem', 'gop')) %>% 
  select(-prez_dem, -prez_gop, -day) %>% 
  arrange(year, month) %>% 
  relocate(year, month)
```


## clean the data in snp.csv 

*   Use separate() to break up the variable date into integer variables year, month, and day.
*   replace month number with month name.
*   convert the 2 digit year to 4 digit year.
*   remove variables: day.
*   arrange according to year and month.
*   make sure year and month are the leading columns.

```{r}
snp = 
  read_csv(file = 'data/fivethirtyeight_datasets/snp.csv') %>% 
  janitor::clean_names() %>% 
  separate(date, c('month', 'day', 'year'), convert = TRUE) %>% 
  mutate(month = month.name[month],
         year = if_else(year >22 , year+1900, year+2000)) %>% 
  select(-day) %>% 
  arrange(year, month) %>% 
  relocate(year, month)
```

## tidy the unemployment data 

*   switching the unemployment data from “wide” to “long” format.
*   replace abbreviated month name to month number and then to month name.
*   arrange according to year and month.
*   make sure year and month are the leading columns.

```{r}
unemployment = 
  read_csv(file = 'data/fivethirtyeight_datasets/unemployment.csv') %>% 
  janitor::clean_names() %>% 
  pivot_longer(jan:dec, 
               names_to = "month",
               values_to = "unemployment_pct") %>% 
  mutate(month = month.name[match(str_to_title(month), month.abb)]) %>% 
  arrange(year, month) %>% 
  relocate(year, month)
```

## Join the datasets 

*   Left join snp into pols_month
*   Left join unemployment into the result.

```{r}
five_thirty_eight = 
  left_join(pols_month, snp, by = c("year", "month")) %>% 
  left_join(unemployment, by = c("year", "month"))
```

## Brief description

### pols-month
The pols-month dataset contains `r ncol(pols_month)` variables and `r nrow(pols_month)` records.

The variables in pols-month dataset include:
`r colnames(pols_month)`

This data set tells us information about national politicians from `r slice(pols_month,1)[1:2]` to `r slice(pols_month,nrow(pols_month))[1:2]`

### snp

The snp dataset contains `r ncol(snp)` variables and `r nrow(snp)` records.

This data set tells us closing values of the S&P stock index from `r slice(snp,1)[1:2]` to `r slice(snp,nrow(snp))[1:2]`

### unemployment

The unemployment dataset contains `r ncol(unemployment)` variables and `r nrow(unemployment)` records.

This data set tells us percentage of unemployment from `r slice(unemployment,1)[1:2]` to `r slice(unemployment,nrow(unemployment))[1:2]`

### resulting dataset

The resulting dataset contains `r ncol(five_thirty_eight)` variables and `r nrow(five_thirty_eight)` records.

The variables include:
`r colnames(five_thirty_eight)`

This data set tells us information about national politicians, closing values of the S&P stock index and percentage of unemployment from `r slice(five_thirty_eight, 1)[1:2]` to `r slice(five_thirty_eight, nrow(five_thirty_eight))[1:2]`

