---
title: "p8105_hw2_yf2605"
author: "Yi_Fang"
date: "`r Sys.Date()`"
output: github_document
---

# Libraries

```{r}
library(tidyverse)
library(readxl)
```

# Problem 1 

## Read and clean the data

```{r}
entrance_exit_raw = 
  read_csv(file = 'data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv',
           col_types = cols(Route8=col_character(), 
                            Route9=col_character(),
                            Route10=col_character(),
                            Route11=col_character(),
                            .default = col_guess()))
entrance_exit_cleaned = 
  janitor::clean_names(entrance_exit_raw) %>% 
  select(line, station_name, station_latitude, 
         station_longitude, starts_with("route"), 
         entry, vending, entrance_type, ada) %>% 
  mutate(entry = recode(entry, "YES" = TRUE, "NO" = FALSE))
```

## Brief description

The dataset contains `r ncol(entrance_exit_raw)` variables: `r colnames(entrance_exit_raw)`. 

Data cleaning steps: 

*   cleaned variable names.
*   retained variables: `r colnames(entrance_exit_cleaned)`. 
*   converted variable `entry` from character to a logical variable, YES-->TRUE, NO-->FALSE.

The size of the resulting dataset is (row x column): `r nrow(entrance_exit_cleaned)` x `r ncol(entrance_exit_cleaned)`. 

These datasets are not tidy.

## Questions

*   How many distinct stations are there?

```{r}
stations = 
  distinct(entrance_exit_cleaned, line, station_name, .keep_all = TRUE)
```

There are `r nrow(stations)` distinct stations.

*   How many stations are ADA compliant?

```{r}
ada_sations = 
  filter(stations, ada == TRUE)
```

`r nrow(ada_sations)` stations are ADA compliant.

*   What proportion of station entrances / exits without vending allow entrance?

```{r}
no_vending_entrance_exit = 
  filter(entrance_exit_cleaned, vending == 'NO')

no_vending_entrance = 
  filter(no_vending_entrance_exit, entry == TRUE)
```

`r round(100 * nrow(no_vending_entrance) / nrow(no_vending_entrance_exit), 2)` % of station entrances / exits without vending allow entrance

*   Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?

```{r}
entrance_exit_tidy = 
  pivot_longer(
    entrance_exit_cleaned, 
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name")

a_train_stations = 
  filter(entrance_exit_tidy, route_name == 'A') %>% 
  distinct(line, station_name, .keep_all = TRUE)

a_train_ada_stations = 
  filter(a_train_stations, ada == TRUE) %>% 
  distinct(line, station_name)
  
```

`r nrow(a_train_stations)` distinct stations serve the A train.

Of the stations that serve the A train, `r nrow(a_train_ada_stations)` are ADA compliant.

# Problem 2

## Read and clean the Mr. Trash Wheel sheet

*   specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
*   use reasonable variable names
*   omit rows that do not include dumpster-specific data
*   round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)

```{r}
mr_trash_wheel = 
  read_excel(path = 'data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx',
             sheet = 'Mr. Trash Wheel',
             range = "A2:N534") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(sports_balls = as.integer(round(sports_balls)))
```

## Read and clean the Professor Trash Wheel sheet

```{r}
professor_trash_wheel = 
  read_excel(path = 'data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx',
             sheet = 'Professor Trash Wheel',
             range = "A2:N116") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(sports_balls = as.integer(round(sports_balls)))
```

## Combine Mr. Trash Wheel and Professor Trash Wheel

```{r}
mr_trash_wheel = 
  mutate(mr_trash_wheel, trash_wheel_type = 'Mr')

professor_trash_wheel = 
  mutate(professor_trash_wheel, trash_wheel_type = 'Professor')

trash_wheels = 
  rbind(mr_trash_wheel, professor_trash_wheel)
```

## Brief description

The dataset contains `r ncol(trash_wheels)` variables and `r nrow(trash_wheels)` records,  `r nrow(mr_trash_wheel)` records are from Mr. Trash Wheel and `r nrow(professor_trash_wheel)` records are from professor_trash_wheel.

The variables include:
`r colnames(trash_wheels)`

The the total weight of trash collected by Professor Trash Wheel is `r sum(professor_trash_wheel$weight_tons)` tons
The total number of sports balls collected by Mr. Trash Wheel in 2020 is `r sum(filter(mr_trash_wheel, year == 2020)$sports_balls)`.

# Problem 3
This problem uses the FiveThirtyEight data; these data were gathered to create the interactive graphic on this page. In particular, we’ll use the data in pols-month.csv, unemployment.csv, and snp.csv. Our goal is to merge these into a single data frame using year and month as keys across datasets.

First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

Join the datasets by merging snp into pols, and merging unemployment into the result.

Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

Note: we could have used a date variable as a key instead of creating year and month keys; doing so would help with some kinds of plotting, and be a more accurate representation of the data. Date formats are tricky, though. For more information check out the lubridate package in the tidyverse.